---
title: "Bayesian MAIHDA for sparse intersections"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian MAIHDA for sparse intersections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  warning = FALSE,
  message = FALSE
)
library(MAIHDA)
library(ggplot2)
# brms needs Stan and a compiler and takes minutes, so it cannot run on CRAN's or
# pkgdown's builders. Its results are precomputed in data-raw/precompute_sparse_vignette.R
# and read from a small cache here; the fast lme4 fits run live.
pc <- readRDS("sparse_precomputed.rds")
pct <- function(x) sprintf("%.0f%%", 100 * x)
```

## Sparse intersectional cells

Intersectional MAIHDA decomposes the between-stratum variation into an **additive**
part (the constituent dimensions' main effects) and an **interaction** part (the
departure from additivity that is genuinely intersectional). A recurring finding in the
literature is that this split is *mostly additive*. That conclusion is only trustworthy,
however, when each intersectional stratum is populated well enough to estimate its
effect.

In practice it often is not. Crossing several dimensions multiplies the strata while the
sample size stays fixed, so cell counts fall away quickly: a few large strata, a long
tail of small ones, and some singletons. Estimating an interaction **variance** from
cells like these is hard, and the two engines fail differently. With `engine = "lme4"`,
maximum likelihood drives the interaction variance to the boundary -- a **singular fit**
-- and reports it as zero with no uncertainty. A real interaction is then read as
*essentially additive*, and nothing in the output flags the estimate as unreliable.

This vignette makes that failure concrete against a **known answer**, and shows what a
Bayesian fit (`engine = "brms"`) does instead. The message is deliberately modest:
weakly-informative priors do **not** recover a point estimate the data cannot support --
at this density no estimator can -- but they replace the singular point with a
**calibrated credible interval** that still admits the true interaction. The interval,
not the point, is what `brms` contributes here.

## A dataset with a known interaction

To judge an estimator we need a dataset whose truth we control. `maihda_sparse_data` is
simulated for exactly that: 4 dimensions form 36 intersectional strata, and the
interaction is constructed to account for **`r pct(pc$truth$gaussian$interaction_share)`
of the between-stratum variance** -- on a Gaussian outcome `y` and on the latent scale of
a binary outcome `event`. Individuals are allocated to strata unevenly, reproducing the
skewed cell sizes of real survey data.

```{r data}
data(maihda_sparse_data)
d <- maihda_sparse_data

cells <- table(interaction(d$gender, d$ethnicity, d$education, d$age_group, drop = TRUE))
summary(as.numeric(cells))                    # cell-size distribution
sum(cells < 5)                                # how many strata have < 5 people
attr(d, "truth")$gaussian$interaction_share   # the true interaction share: 0.40
```

A third of the strata hold fewer than five individuals. This is not an edge case -- it
is the usual situation once more than two or three dimensions are crossed -- and it is
precisely where maximum likelihood becomes unreliable.

## What lme4 reports

We fit the crossed-dimensions decomposition with `engine = "lme4"` on both outcomes.

```{r lme4}
m_g <- maihda(y ~ 1 + (1 | gender:ethnicity:education:age_group),
              data = d, decomposition = "crossed-dimensions", engine = "lme4")
m_b <- maihda(event ~ 1 + (1 | gender:ethnicity:education:age_group),
              data = d, decomposition = "crossed-dimensions",
              engine = "lme4", family = "binomial")

c(gaussian = m_g$decomposition$interaction_share,
  binary   = m_b$decomposition$interaction_share)
c(gaussian_singular = isTRUE(m_g$model$diagnostics$singular),
  binary_singular   = isTRUE(m_b$model$diagnostics$singular))
```

Both fits are **singular**: maximum likelihood has placed the interaction variance at
the boundary. The Gaussian interaction share is pulled down to
`r pct(pc$lme4$gaussian$share)`, and the binary share to `r pct(pc$lme4$binary$share)` --
against a true 40%, the binary fit reports a substantial interaction as effectively
absent. Crucially, neither estimate carries an interval. The estimate is not only biased
but reported with unwarranted precision, because the model has no way to express how
little the sparse cells determine.

## What brms adds: a calibrated interval

Refitting with `engine = "brms"` and a weakly-informative prior on the random-effect
standard deviations changes the output in one important respect. The prior regularises
the between-stratum standard deviations away from the boundary -- so the fit is not
singular -- and the posterior yields a credible interval in place of a single number.
The call is shown but not evaluated here (`brms` requires Stan and several minutes; see
the note on precomputation below).

```{r brms-call, eval = FALSE}
m_g_brms <- maihda(
  y ~ 1 + (1 | gender:ethnicity:education:age_group),
  data = d, decomposition = "crossed-dimensions", engine = "brms",
  prior = brms::set_prior("normal(0, 0.5)", class = "sd"),
  chains = 4, iter = 2000, warmup = 1000, cores = 4, seed = 1,
  control = list(adapt_delta = 0.97)
)
m_g_brms$decomposition$interaction_share
m_g_brms$decomposition$interaction_share_ci
```

The cached posterior summaries:

```{r brms-summary}
bg <- pc$brms$gaussian; bb <- pc$brms$binary
data.frame(
  outcome      = c("Gaussian", "Binary"),
  true_share   = c(pc$truth$gaussian$interaction_share, pc$truth$binary_latent$interaction_share),
  brms_share   = c(bg$share, bb$share),
  ci_low       = c(bg$ci[1], bb$ci[1]),
  ci_high      = c(bg$ci[2], bb$ci[2]),
  max_rhat     = c(bg$diag$max_rhat, bb$diag$max_rhat),
  divergences  = c(bg$diag$divergences, bb$diag$divergences)
)
```

The posterior **point** remains low: the interaction is weakly identified at this
density, so `brms` does not recover the 40% either, and we should not expect it to. The
**credible interval**, however, runs from near zero to past the true value, correctly
conveying that the data are consistent with an interaction anywhere in that range. This
is the substantive difference. Maximum likelihood reported a singular point as if it
were certain; `brms` reports an interval that *includes the truth* and makes the
uncertainty explicit. The diagnostics (`max_rhat` close to 1, no divergent transitions)
confirm the posterior is well-sampled.

## Comparison

```{r fig-share, fig.height = 3.8}
fig <- data.frame(
  outcome = factor(rep(c("Gaussian", "Binary"), each = 2), levels = c("Gaussian", "Binary")),
  method  = rep(c("lme4 (ML)", "brms (Bayesian)"), 2),
  share   = c(pc$lme4$gaussian$share, pc$brms$gaussian$share,
              pc$lme4$binary$share,   pc$brms$binary$share),
  lo      = c(NA, pc$brms$gaussian$ci[1], NA, pc$brms$binary$ci[1]),
  hi      = c(NA, pc$brms$gaussian$ci[2], NA, pc$brms$binary$ci[2])
)
truth <- pc$truth$gaussian$interaction_share

ggplot(fig, aes(method, share, colour = method)) +
  geom_hline(yintercept = truth, linetype = "dashed") +
  geom_pointrange(aes(ymin = lo, ymax = hi), na.rm = TRUE, linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ outcome) +
  scale_y_continuous("Interaction share of between-stratum variance",
                     labels = function(x) paste0(round(100 * x), "%")) +
  labs(x = NULL, colour = NULL,
       title = "A singular ML point estimate vs. a calibrated brms interval",
       subtitle = "Dashed line: true interaction share (40%)") +
  theme_minimal() + theme(legend.position = "none")
```

Both point estimates fall below the true value; neither engine can locate an interaction
the sparse data do not pin down. They differ in what is reported alongside the point. The
`brms` credible interval spans the true 40%, whereas `lme4` provides a single value with
no uncertainty. Reading the `lme4` point at face value -- an interaction of roughly
3--23%, or none at all -- is the error the singular fit invites.

## Why the interval, not the point, is what changes

The per-stratum effects (the stratum random intercepts) are similar under the two
engines: both apply partial pooling, shrinking thinly-estimated strata toward the
additive prediction. What differs is the treatment of uncertainty. Maximum likelihood
commits to a single variance decomposition and, at the boundary, reports it without a
standard error. `brms` propagates the full posterior, so the interaction **share** is
accompanied by an interval whose width reflects how little the sparse cells constrain it
-- here, wide enough to span everything from near zero to past the true 40%. Summarising
the posterior by its point discards exactly the information that matters: the degree of
uncertainty.

## Practical guidance

* **Use a mildly informative prior on `class = "sd"`.** A half-normal (`normal(0, s)`,
  used here), `exponential(rate)`, or `student_t(3, 0, s)` scaled to the outcome
  (approximately the residual SD for a Gaussian outcome; about 1 on the logit scale for a
  binary one) regularises the between-stratum standard deviations away from the boundary.
  Wider, near-flat priors are not preferable here: under heavy sparsity `brms`'s default
  and `student_t(3, 0, 1)` produced divergent transitions and an over-confident,
  near-zero interaction, whereas `normal(0, 0.5)` sampled cleanly and produced the
  interval that covered the truth. Target a specific term with the `group =`/`coef =`
  arguments of `brms::set_prior()`.
* **Check the sampler.** Sparse cells produce a difficult posterior geometry. Raise
  `adapt_delta` (0.95--0.99) to remove divergent transitions, and confirm `Rhat` and the
  effective sample size before interpreting the interval.
* **Report the interval, not the point.** With this little data per cell the point
  estimate is uninformative; the calibrated interval is the appropriate summary, and it
  is exactly what the singular `lme4` fit cannot provide.
* **Precompute brms for vignettes and reports.** `brms` requires Stan and is slow, so it
  should not run at build time. Fit it once in a script
  (`data-raw/precompute_sparse_vignette.R`), save the small summaries needed (not the
  multi-megabyte `brmsfit` object), and render from the cache, as this vignette does.

## Caveats

This is a **simulated** dataset with a constructed interaction, which is what allows the
estimators to be judged against a known value; real data provide no such benchmark, and
the *mostly additive* pattern reported across the MAIHDA literature is frequently
genuine. The conclusion is correspondingly narrow: when intersectional cells are sparse,
a singular maximum-likelihood interaction of zero is not evidence that the interaction is
zero, and a regularised Bayesian fit with an explicit interval is the more defensible
summary.
