---
title: "MAIHDA for Binary Outcomes (Discriminatory Accuracy)"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MAIHDA for Binary Outcomes (Discriminatory Accuracy)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Why binary outcomes?

Many of the original MAIHDA applications target a **binary** health outcome --
obese vs. not, hypertensive vs. not, screened vs. not. Merlo's framing of
*discriminatory accuracy* is, at heart, a question about a binary classifier: how
well do the intersectional strata alone separate people who do and do not have the
outcome?

For a binary outcome the model is a multilevel **logistic** regression:

\[
\text{logit}\,\Pr(y_{ij} = 1) = \beta_0 + u_j, \qquad u_j \sim N(0, \sigma_u^2),
\]

where \(j\) indexes the intersectional stratum. MAIHDA reads off two
complementary quantities from this model:

- the **VPC/ICC**, the share of the (latent) variation that lies *between* strata;
- the **discriminatory accuracy** (e.g. the AUC / C-statistic), how well stratum
  membership predicts the individual outcome.

A high between-stratum VPC can still go with only *moderate* discriminatory
accuracy at the individual level -- that contrast is the whole point of the "DA"
in MAIHDA, and this vignette shows how to get both numbers.

```{r load}
library(MAIHDA)
data("maihda_health_data")

# A two-level outcome: obese (Yes) vs. not (No)
table(maihda_health_data$Obese)
```

## Fitting a logistic MAIHDA model

`fit_maihda()` detects a two-level outcome automatically. If you leave `family` at
its default, the function switches to `family = "binomial"`, recodes the response
to 0/1 for `glmer()`, and warns you so the change is never silent:

```{r autodetect}
model_null <- fit_maihda(
  Obese ~ 1 + (1 | Gender:Race:Education),
  data = maihda_health_data
)
```

The warning is a feature, not a problem. To be explicit (and to silence the
warning), pass `family = "binomial"` yourself -- this is the recommended form for
scripts:

```{r explicit, eval=FALSE}
model_null <- fit_maihda(
  Obese ~ 1 + (1 | Gender:Race:Education),
  data   = maihda_health_data,
  family = "binomial"
)
```

> If you actually want a linear probability model on a 0/1 outcome, ask for it
> explicitly with `family = "gaussian"`; the auto-switch only fires for the
> default family.

## The VPC is on the latent scale

`summary()` reports the VPC/ICC for the logistic model. There is no observed-scale
residual variance for a Bernoulli outcome, so MAIHDA uses the standard
**latent-scale** approximation: the level-1 (within-stratum) variance is fixed at
\(\pi^2/3 \approx 3.29\) for the logit link (and the corresponding constant for a
probit link). The VPC is therefore the between-stratum share of variation on that
underlying latent scale, not on the probability scale.

```{r summary-null}
summary_null <- summary(model_null)
print(summary_null)
```

Read from this null model, the VPC is the **total** between-stratum share of
latent variation in the odds of obesity. As in the Gaussian case, the stratum
random effects capture the *combined* additive + interaction differences across
strata; they isolate the *pure* intersectional (interaction) component only once
the additive main effects of the strata variables are in the model.

> For a binomial model `summary()` now reports the discriminatory accuracy (AUC /
> Median Odds Ratio) automatically, so the printed summary above already carries that
> block -- the [Discriminatory accuracy](#discriminatory-accuracy-auc-and-median-odds-ratio)
> section below explains how to read it and how to obtain the pieces on their own.

For an interpretable **probability-scale** complement to the latent-scale VPC, the
package provides `maihda_vpc_response()`, which estimates the response-scale VPC by
simulation (Goldstein, Browne & Rasbash 2002):

```{r vpc-response}
maihda_vpc_response(model_null, seed = 1)
```

Report it *alongside* -- not instead of -- the latent-scale VPC: it depends on the
overall prevalence, and for adjusted models it is evaluated at the mean covariate
profile, so it is best read from the null model.

## Adjusted model and PCV

A "Model 2" adds individual-level covariates to ask how much of the
between-stratum variation they account for. Here we adjust for age, fit on the
**same analytic sample and strata** as the null model, and compare with
`calculate_pcv()`. PCV compares variances across models, so both must use the same
complete-case sample:

```{r adjusted}
health_complete <- maihda_health_data[complete.cases(
  maihda_health_data[, c("Obese", "Age", "Gender", "Race", "Education")]
), ]

model_null2 <- fit_maihda(
  Obese ~ 1 + (1 | Gender:Race:Education),
  data = health_complete, family = "binomial"
)

# Model 2: adjust for an individual-level covariate (age)
model_adj <- fit_maihda(
  Obese ~ Age + (1 | Gender:Race:Education),
  data = health_complete, family = "binomial"
)

pcv <- calculate_pcv(model_null2, model_adj)
print(pcv)
```

The same caveats as in the continuous case apply, with one extra wrinkle: for
non-Gaussian models the latent residual variance is fixed by the link, so a change
in the between-stratum variance is partly a **rescaling** of the latent scale, not
only "variance explained". Interpret the PCV as a model-dependent, descriptive
change, not a causal decomposition.

> **A note on adjusting for the strata's own categories.** If instead you add the
> *categorical* main effects that define the strata (e.g.
> `Obese ~ Gender + Race + Education + ...`) to recover the additive-vs-interaction
> split shown in the [introduction](introduction.html), be aware that in a logistic
> model those fixed effects are nearly collinear with the stratum random intercept,
> so `glmer()` often reports a convergence or "nearly unidentifiable" note. Scaling
> covariates and using `control = lme4::glmerControl(optimizer = "bobyqa")` usually
> helps.

## What "interaction" means on the logit scale

Once those main effects *are* in the model, the stratum random effect is the pure
intersectional component, and `maihda_interactions()` flags the strata whose
component is credibly non-zero:

```{r interactions}
model_int <- fit_maihda(
  Obese ~ Gender + Race + Education + (1 | Gender:Race:Education),
  data = health_complete, family = "binomial"
)

maihda_interactions(model_int)
```

Read the numbers on the scale they were computed on. In the Gaussian case the
stratum BLUP \(u_j\) is the part of the stratum's mean outcome attributable to
interaction, in the outcome's own units. Here it is a deviation in **log-odds** --
so what is flagged is *multiplicative*, odds-scale interaction. Evans et al. (2024,
sec. 2.5.1) say it plainly: in a logistic MAIHDA you can "no longer directly
interpret \(u_j\) as the change in mean outcomes (i.e., shift in probabilities)
attributable to interaction effects". The printed output says so too, which is why
the binomial fit above carries a line the Gaussian one does not -- and why it prints
a `prob_diff` column the Gaussian one does not: the same departure expressed in
probability points, so the log-odds figure is never the only number on screen. That
column is a display convenience; `scale = "response"` below is how to get it, with
its interval, as data.

There are in fact **three** quantities in play here, and this analysis reports two of
them. Keeping them apart is the whole of the difficulty.

**1. The multiplicative interaction, \(u_j\)** -- whether a dimension multiplies the
odds by the same factor at every level of the others. That is what was just printed,
and what `flagged` is about.

**2. The same departure in probability points, \(\pi^{B}_{j}\)** -- the units to
report if your reader thinks in risks. More on it in a moment.

**3. The additive (risk-difference) interaction** -- whether a dimension adds the
same number of *percentage points* of risk at every level of the others. Neither 1
nor 2 reports this, and it is generally non-zero even when \(u_j\) is exactly zero.
Take a model with a \(-2\) baseline and \(+0.7\) for each of two dimensions and no
interaction whatsoever:

```{r scale-dependence}
p <- function(a, b) plogis(-2 + 0.7 * a + 0.7 * b)

c(`risk added by B, at A = 0` = p(0, 1) - p(0, 0),
  `risk added by B, at A = 1` = p(1, 1) - p(1, 0))
```

The odds ratio for B is identical at both levels of A -- that is what \(u_j = 0\)
means -- yet B adds nine and a half points of risk in one group and fourteen in the
other. The excess is a property of the logistic curve, which is steeper in the middle
than in the tails, not a finding about the world. So **"no strata flagged" supports
"no credible multiplicative interaction", not "no interaction"**: quantity 3 is
generally non-zero regardless, and it is often the one a policy audience cares about.

*Generally*, not always -- worth knowing, because the exception is a perfectly
ordinary design rather than a curiosity. The risk-difference interaction is positive
below the curve's midpoint and negative above it, so it must pass through zero, and
it does so where the two comparisons straddle the midpoint symmetrically:

```{r scale-dependence-zero}
rd_int <- function(b0, a, b) {
  (plogis(b0 + a + b) - plogis(b0 + a)) - (plogis(b0 + b) - plogis(b0))
}

vapply(c(`-2.0` = -2, `-0.7` = -0.7, `0.0` = 0), rd_int, numeric(1), a = 0.7, b = 0.7)
```

At an intercept of \(-(a + b)/2 = -0.7\) the risks add exactly. That is one
configuration out of many, and nothing in the flags tells you which one you are in --
which is the point: they are not evidence about quantity 3 in either direction.

Quantity 2, then. Evans et al. (2024) define it as the gap between a stratum's total
predicted probability and the probability implied by the additive main effects
alone,

\[
\pi^{B}_{j} = \pi_j - \pi^{A}_{j}, \qquad
\pi_j = \mathrm{logit}^{-1}(x_j'\beta + u_j), \qquad
\pi^{A}_{j} = \mathrm{logit}^{-1}(x_j'\beta),
\]

and rank-plot \(\pi^{B}_{j}\) where the linear case plots \(u_j\). Ask for it with
`scale = "response"`:

```{r interactions-response}
maihda_interactions(model_int, scale = "response")
```

Same strata, same evidence, readable units: a flagged stratum's interaction is now a
difference in **percentage points** of predicted probability rather than a log-odds
departure.

That correspondence is exact, not approximate. Write \(g\) for the map from a
stratum's BLUP to its \(\pi^{B}_{j}\); \(g\) is strictly increasing and
\(g(0) = 0\), so the estimate, both interval endpoints and zero carry across
together. `flagged`, `direction` and the p-values are identical under either scale,
and the response-scale interval is the exact image of the link-scale one -- no
simulation, and no delta-method approximation. It inherits the same conditionality
(fixed effects and variance components held at their point estimates). Two things do
change: `se` is dropped, because the resulting interval is deliberately asymmetric
about the estimate and no single standard error would reproduce it; and the ranking
can differ, since the same log-odds departure buys more probability near
\(\pi = 0.5\) than out in the tail.

A `rope` is then read in the same units, which is often where the argument becomes
concrete -- "are more than two percentage points of this stratum's risk attributable
to its interaction?" is a question a reader can actually answer:

```{r interactions-rope}
table(maihda_interactions(model_int, scale = "response", rope = 0.02)$decision)
```

**If it is quantity 3 you want**, neither scale will give it to you -- `scale =
"response"` re-expresses the multiplicative interaction, it does not replace it with
the risk-difference one. Fit a linear probability model instead: `family = "gaussian"`
on the 0/1 outcome, the option `fit_maihda()` points at when it auto-detects a binary
outcome. There the BLUP *is* the risk-difference interaction by construction, at the
usual costs (predictions outside \([0, 1]\), heteroskedastic residuals).

## Discriminatory accuracy (AUC and Median Odds Ratio)

The VPC summarises *variation*; discriminatory accuracy summarises *prediction*.
For a binomial model this is reported **automatically**: `summary()` of a binomial
`maihda_model` carries a `discriminatory_accuracy` slot, and `maihda(...,
family = "binomial")` surfaces it on its summaries and headline `print()`. The
explicit `maihda_discriminatory_accuracy()` below is the same quantity on its own --
it bundles the two individual-level summaries for a logistic MAIHDA model: the
**AUC / C-statistic** (how well the predicted probabilities separate cases from
non-cases) and the **Median Odds Ratio (MOR)** (the between-stratum heterogeneity
expressed on the odds-ratio scale). The strata-only model's AUC is the
discriminatory accuracy of the intersectional strata *themselves* -- Merlo's central
quantity. Comparing it with the adjusted model shows whether individual information
beyond stratum membership sharpens classification:

```{r da}
da_null <- maihda_discriminatory_accuracy(model_null2)
da_adj  <- maihda_discriminatory_accuracy(model_adj)

da_null
da_adj
```

You can also call the pieces directly: `maihda_auc(prob, y)` on any vector of
predicted probabilities and 0/1 outcomes (it equals the Mann-Whitney U statistic,
so it needs no extra package), and `maihda_mor(model)` for the Median Odds Ratio.

```{r auc-direct}
prob_null <- predict_maihda(model_null2, type = "individual", scale = "response")
y_obs     <- as.numeric(lme4::getME(model_null2$model, "y"))

maihda_auc(prob_null, y_obs)
```

An AUC of 0.5 is chance. Here both models sit around 0.6: even a non-trivial
between-stratum VPC translates into only **modest** accuracy at the *individual*
level -- exactly the cautionary message that motivates discriminatory-accuracy
reporting in MAIHDA. Note the adjusted model barely moves the AUC: the
categorical covariates that *define* the strata are already captured by stratum
membership, so only the continuous covariate (`Age`) adds genuinely new
individual-level information.

> **The MOR needs the logit link.** The Median Odds Ratio is an odds-ratio-scale
> quantity, so it is defined only for `binomial(link = "logit")` (the default). For a
> `binomial(link = "probit")` fit, `maihda_mor()` errors and
> `maihda_discriminatory_accuracy()` reports the AUC with `mor = NA` -- the AUC,
> being rank-based, is link-agnostic.

## Plots adapt to the binomial family

The standard plots recognise the binomial family and switch to the probability
scale and to deviance-based diagnostics automatically:

```{r plot-predicted}
# Predicted probabilities per stratum with intervals
plot(model_adj, type = "predicted")
```

```{r plot-vpc}
# Latent-scale variance partition
plot(model_adj, type = "vpc")
```

```{r plot-prediction-deviation, warning = FALSE}
# For binomial fits the dashboard highlights the largest absolute
# deviance residuals rather than raw deviations from the mean.
plot(model_adj, type = "prediction_deviation")
```

See the [plot interpretation vignette](interpreting_plots.html) for how to read
each of these.

## Count outcomes work the same way

A Poisson (count) outcome follows the identical pattern -- pass
`family = "poisson"` to `fit_maihda()`. The VPC then uses the log-link
latent-scale residual variance, and the summary, PCV, and plotting helpers all
behave as above.

One thing to watch with counts: that latent-scale residual variance is an
*approximation*, and which one you use matters when the outcome is rare.
`fit_maihda(count_approximation = )` offers the three of Nakagawa, Johnson &
Schielzeth (2017) -- `"lognormal"` (the default), `"delta"` and `"trigamma"`.
They agree above a mean count of about 2 and diverge sharply below it: at a mean
count of 0.34 they give level-1 variances of 1.37, 2.94 and 9.76, so the same
model reports a VPC six times larger under one than another. `summary()`
therefore names the approximation and the mean count it was evaluated at, and
warns when that count is at or below 2. Report which one you used.

## References

- Merlo, J. (2018). Multilevel analysis of individual heterogeneity and discriminatory accuracy (MAIHDA) within an intersectional framework. *Social Science & Medicine*, 203, 74-80.

- Evans, C. R., Williams, D. R., Onnela, J. P., & Subramanian, S. V. (2018). A multilevel approach to modeling health inequalities at the intersection of multiple social identities. *Social Science & Medicine*, 203, 64-73.

- Evans, C. R., Leckie, G., Subramanian, S. V., Bell, A., & Merlo, J. (2024). A tutorial for conducting intersectional multilevel analysis of individual heterogeneity and discriminatory accuracy (MAIHDA). *SSM - Population Health*, 26, 101664. doi:10.1016/j.ssmph.2024.101664

- Merlo, J., Wagner, P., Ghith, N., & Leckie, G. (2016). An original stepwise multilevel logistic regression analysis of discriminatory accuracy: the case of neighbourhoods and health. *PLOS ONE*, 11(4), e0153778.
