---
title: "Cross-classified MAIHDA: additive vs. interaction in one model"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-classified MAIHDA: additive vs. interaction in one model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Two ways to split intersectional inequality

Intersectional MAIHDA asks how much of the variation in an outcome lies *between*
intersectional strata, and how much of that between-stratum variation is **additive**
(explained by the constituent dimensions' main effects) versus a genuinely
**intersectional interaction** (over and above the additive parts).

The package offers two estimators for that split, selected with the `decomposition`
argument of `maihda()`:

* **`"two-model"` (default)** -- the classic approach. Fit a *null* model and an
  *adjusted* model that adds the dimensions' additive main effects as **fixed**
  effects, and read the additive share from the **PCV** (the proportional change in
  between-stratum variance). See `vignette("introduction", package = "MAIHDA")`.
* **`"cross-classified"`** -- a **single** model that enters each dimension's additive
  main effect as a **random** intercept:

  $$y_i = \beta_0 + \mathbf{x}_i\boldsymbol\beta
    + u^{(1)}_{d_1[i]} + \dots + u^{(K)}_{d_K[i]} + u^{(\text{stratum})}_{s[i]} + e_i.$$

  Each dimension's random-effect variance is its additive contribution, and the full
  intersection (`stratum`) random-effect variance is the interaction *beyond* additive.
  The additive and interaction shares of the total between-strata variance are read
  directly from this one fit.

This is exactly the cross-classified random-effects design used in the published
cross-classified MAIHDA studies. It is fit with ordinary multilevel software --
`lme4` for a frequentist fit or `brms` for a Bayesian one -- so it does **not** require
any special toolchain.

## Running a cross-classified analysis

```{r fit}
library(MAIHDA)
data("maihda_health_data")

cc <- maihda(
  BMI ~ Age + (1 | Gender:Race:Education),
  data = maihda_health_data,
  decomposition = "cross-classified"
)
cc
```

`maihda()` builds the cross-classified model for you from the intersectional shorthand:
it reads the dimensions (`Gender`, `Race`, `Education`) from the grouping term, adds one
additive random intercept per dimension plus the intersection random intercept, and
fits the single model:

```{r formula}
cc$formula
```

The partition is on `cc$decomposition` (and printed above):

```{r decomposition}
cc$decomposition$additive_var        # sum of the dimension random-effect variances
cc$decomposition$interaction_var     # the intersection random-effect variance
cc$decomposition$additive_share      # additive share of the between-strata variance
cc$decomposition$interaction_share   # the complement: the interaction share
cc$decomposition$per_dim             # additive variance per dimension
```

`summary()` shows the full variance-components table (one row per dimension, the
interaction, and the residual) alongside the decomposition:

```{r summary}
summary(cc$model)
```

### Figures

The variance-partition and decomposition figures are cross-classified-aware: the VPC
plot shows one additive slice per dimension plus the interaction and residual, and the
deviation decomposition splits each stratum's deviation into its additive
(dimension random effects) and interaction (intersection random effect) parts.

```{r plots, eval = FALSE}
plot(cc, type = "vpc")            # per-dimension additive + interaction + residual
plot(cc, type = "effect_decomp")  # additive vs. interaction, per stratum
plot(cc, type = "ternary")        # additive / interaction / uncertainty per stratum
```

## Comparing across a higher-level group

Pass a `group` to decompose within each level of a higher-level variable -- here, how
the additive-vs-interaction split differs across countries in the PISA data:

```{r groups, eval = FALSE}
data("maihda_country_data")
cc_grp <- maihda(
  math ~ 1 + (1 | gender:ses),
  data = maihda_country_data,
  group = "country",
  decomposition = "cross-classified"
)
plot(cc_grp, type = "group_additive_share")  # additive share by country
plot(cc_grp, type = "group_components")       # additive / interaction / residual
```

## A Bayesian fit

Set `engine = "brms"` for a Bayesian cross-classified fit; the additive and interaction
shares then carry posterior **credible** intervals (no bootstrap needed). This is the
recommended engine when dimensions have few levels (see the caveats below).

```{r brms, eval = FALSE}
cc_b <- maihda(
  BMI ~ Age + (1 | Gender:Race:Education),
  data = maihda_health_data,
  engine = "brms",
  decomposition = "cross-classified"
)
cc_b$decomposition$additive_share_ci
```

## Two important caveats

1. **The additive share is not the PCV.** The two-model PCV and the cross-classified
   additive share target the same idea -- how much of the between-strata variance is
   additive -- but with **different estimators** (fixed main effects across two models
   vs. a single model's random main-effect variances, which are partially pooled). They
   will be close but not numerically identical; do not treat one as a check on the
   other.

2. **Few-level dimensions are poorly identified.** A dimension's additive variance is
   estimated from its handful of levels. A binary dimension (e.g. a two-level sex
   variable) contributes a variance estimated from just two groups, which `lme4` will
   often pin at the boundary (a *singular fit*), inflating the additive share and
   shrinking the interaction toward zero. Watch for the singular-fit note in the
   output, and prefer `engine = "brms"`, whose weakly-informative priors regularise
   these variances, when dimensions are few-levelled.

## Genuinely cross-classified *data*

The same machinery also serves a genuinely cross-classified data structure -- where
units belong to two non-nested higher-level groupings (for example strata crossed with
region, or an age--period--cohort design). Build the model with `fit_maihda()` using a
`stratum` column from `make_strata()` plus the extra grouping factor, e.g.
`outcome ~ covars + (1 | stratum) + (1 | region)`, and `summary()` reports each
grouping factor's variance component. The `decomposition = "cross-classified"` mode of
`maihda()` is the *additive/interaction* special case of this, where the extra grouping
factors are the strata's own dimensions.
