---
title: "Case study: youth unemployment across Europe (ESS)"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: youth unemployment across Europe (ESS)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE)
library(MAIHDA)
# This vignette uses real European Social Survey microdata, which is licence-
# restricted and therefore NOT shipped with the package. Nothing is fitted at build
# time; the analysis is reproduced from a small cache written by
# data-raw/precompute_youth_vignette.R. The model calls below are shown but not run.
pc <- readRDS("youth_precomputed.rds")
pct <- function(x) sprintf("%.1f%%", 100 * x)
pf  <- function(p) if (p < 0.001) "< 0.001" else sprintf("%.3f", p)
m <- pc$meta; U <- pc$unemp; N <- pc$neet
fig <- function(f) knitr::include_graphics(file.path("figures", f))
```

## What this case study does

The other vignettes explain MAIHDA's machinery on data chosen to teach it. This one
goes the other way: it takes a real, contested empirical question -- *how
intersectional is youth unemployment in Europe?* -- and walks the full MAIHDA
workflow on real survey data, reporting what it finds, including where the honest
answer is "we cannot tell". It leans on, rather than repeats, the method vignettes:
see `vignette("cross_classified")` for the contextual design used here,
`vignette("binary_outcomes")` for discriminatory accuracy, and
`vignette("bayesian_sparse_maihda")` for the singular-fit problem that returns at
the end.

The headline is deliberately undramatic and is the point of the exercise:
**MAIHDA reveals a strong intersectional _gradient_ in youth unemployment, finds it
essentially _additive_ rather than multiplicative, and -- on a harder outcome --
shows why a single-number interaction estimate should not be trusted.** Additivity is
not the absence of intersectionality; it is an empirical result about its *form*, and
it is the form most MAIHDA studies report.

## Data

The data are the **European Social Survey**, integrated files for **Rounds `r m$rounds`**,
pooled. The analysis sample is **`r format(m$n_unemp, big.mark = ",")` economically
active young people aged 15--29** across **`r m$n_countries` countries**. "Economically
active" is the ILO labour-force base -- in paid work, or unemployed and actively
looking -- which is the youth-unemployment-rate denominator and which excludes
students, so *completed* education is meaningful. The pooled unemployment rate is
**`r pct(m$unemp_rate)`**.

The intersectional strata cross three dimensions into **`r m$n_strata` strata**:

* **gender** (male / female),
* **migration background** (native / second-generation / first-generation, from the
  respondent's and parents' country of birth),
* **education** (low / medium / high, ES-ISCED collapsed).

Every stratum holds at least `r m$min_cell_unemp` people, so the between-stratum
variance is well estimated -- this is *not* the sparse regime of
`vignette("bayesian_sparse_maihda")`.

```{r build, eval = FALSE}
# The ESS microdata is licence-restricted; download the integrated .dta files into
# data-raw/ess/ (see data-raw/ess/README.md) and run data-raw/fetch_ess_youth.R,
# which builds the analysis frame `ess_youth` used below.
```

## The design: country as context, not stratum

Youth unemployment ranges from under 3% to nearly 40% *between* countries, so country
cannot be ignored. But country-by-stratum cells are thin, so folding country *into* the
strata would force the analysis into the sparse regime. The contextual
cross-classified MAIHDA solves this: country enters as a *crossed* random intercept,
estimated from `r m$n_countries` well-populated levels, while the intersectional
strata stay clean.

```{r fit, eval = FALSE}
au <- maihda(unemployed ~ gender + migration + education + (1 | gender:migration:education),
             data = ess_youth, context = "country", family = "binomial",
             response_vpc = TRUE)
```

## What MAIHDA reveals: a real intersectional gradient

```{r partition, echo = FALSE, out.width = "100%"}
fig("ess_partition.png")
```

The variance partition puts **`r pct(U$vpc_stratum)`** of the (latent-scale) variation
*between intersectional strata* and **`r pct(U$vpc_country)`** between countries -- the
contextual effect is about twice the intersectional one. The strata are far from inert,
though: the discriminatory accuracy is **AUC = `r sprintf("%.3f", U$auc)`** with a
**median odds ratio of `r sprintf("%.2f", U$mor)`** (`r format(U$cases, big.mark=",")`
unemployed vs. `r format(U$controls, big.mark=",")` employed). An AUC of
`r sprintf("%.2f", U$auc)` is high by MAIHDA standards: knowing only a young person's
gender, migration background and education separates the unemployed from the employed
fairly well. That separation -- a 3--4&times; spread in risk across strata -- *is* the
intersectional inequality, and MAIHDA has surfaced it.

```{r country, echo = FALSE, out.width = "85%"}
fig("ess_country.png")
```

## Additive or multiplicative?

The substantive intersectionality question is whether the multiply-disadvantaged are
worse off than the *sum* of their disadvantages -- a genuine interaction -- or merely
the sum of them. The two-model PCV answers it: the additive main effects of gender,
migration and education explain **`r pct(U$pcv$value)`** of the between-stratum variance.

```{r effectdecomp, echo = FALSE, out.width = "100%"}
fig("ess_effect_decomp.png")
```

A PCV at the ceiling, with a singular adjusted fit, *could* be the over-shrinkage
artefact of `vignette("bayesian_sparse_maihda")`. Here it is not -- it is corroborated
two ways that do not depend on the random-effect machinery:

* a shrinkage-free likelihood-ratio test of the full three-way interaction against the
  additive model is non-significant (&chi;&sup2; = `r sprintf("%.1f", U$fe$chisq)`,
  `r U$fe$df` df, *p* = `r pf(U$fe$p)`; AIC prefers the additive model), and
* the additive model reproduces every one of the `r m$n_strata` observed stratum rates
  to within **`r sprintf("%.1f", 100 * U$maxresid)` percentage points**:

```{r additivity, echo = FALSE, out.width = "85%"}
fig("ess_additivity.png")
```

So a low-educated migrant woman is at high risk -- but almost exactly the risk that
adding "female", "migrant" and "low-educated" predicts, with no extra multiplicative
penalty. Education dominates the gradient; gender and migration add modestly. This
*mostly-additive* pattern is the rule across the MAIHDA literature, not an anomaly.

## A harder probe: NEET and the migrant-women hypothesis

Unemployment-among-the-active excludes the inactive, and intersectionality theory
locates one of its clearest predictions there: young migrant women drawn into
home/care roles. The **NEET** indicator (not in employment, education or training),
measured on **all** `r format(m$n_neet, big.mark = ",")` young people (rate
`r pct(m$neet_rate)`), pulls that channel back in. The raw gender-by-migration NEET
rates show exactly the predicted shape:

```{r neetcell, echo = FALSE}
ct <- N$cell
tab <- tapply(ct$rate, list(ct$gender, ct$migration), function(x) sprintf("%.1f%%", 100 * x))
knitr::kable(tab[, c("native", "second_gen", "first_gen")], caption = "NEET rate by gender and migration background")
```

Migration barely moves NEET for men or for native/second-generation women -- and then
first-generation migrant women jump. A joint likelihood-ratio test of the two-way
interactions is borderline-significant (&chi;&sup2; = `r sprintf("%.1f", N$fe2$chisq)`,
`r N$fe2$df` df, *p* = `r pf(N$fe2$p)`). There is, at least, a hint of a genuine
interaction here.

## The trap, and the honest answer

What does the MAIHDA decomposition make of that hint? With `engine = "lme4"`, nothing:
the adjusted NEET model is **singular**, the interaction variance pinned to exactly
zero, PCV = `r pct(N$lme4$pcv)` -- the hint vanishes, reported as certain additivity.
With only `r m$n_strata` strata, an interaction *variance* this small is at the edge of
what maximum likelihood can estimate, and it fails silently. This is the same failure
as `vignette("bayesian_sparse_maihda")`, now on real data.

A Bayesian refit with a mildly-informative prior is the fix. It does not manufacture an
effect the data cannot support; it replaces the singular point with an honest interval.

```{r brms-call, eval = FALSE}
# Precomputed in data-raw/precompute_youth_vignette.R (brms needs Stan and is slow).
f_adj <- brm(neet ~ gender + migration + education + (1 | stratum) + (1 | country),
             family = bernoulli(), data = ess_youth_neet,
             prior = c(prior(normal(0, 1), class = "b"),
                       prior(normal(0, 0.5), class = "sd")),
             chains = 4, iter = 4000, warmup = 1500, control = list(adapt_delta = 0.95))
```

```{r brms-fig, echo = FALSE, out.width = "75%", eval = !is.null(N$brms)}
fig("ess_neet_brms.png")
```

```{r brms-text, echo = FALSE, results = "asis", eval = !is.null(N$brms)}
b <- N$brms; mw <- b$interaction[grepl("first", b$interaction$term), ]
cat(sprintf(
"Two things change. The adjusted between-stratum interaction SD is no longer a singular zero but a credible interval, **%.3f, 95%% CrI [%.3f, %.3f]** -- small, with a lower bound near the floor. And the explicit migrant-women interaction carries an odds ratio of **%.2f** with a **%.0f%%** posterior probability of being a penalty, but a 95%% credible interval (**%.2f--%.2f**) that still spans 1. (Sampler: max R-hat %.3f, %d divergences.)\n\n",
  b$sd_stratum$med, b$sd_stratum$lo, b$sd_stratum$hi,
  mw$or_med, 100 * mw$p_gt0, mw$or_lo, mw$or_hi, b$diag$max_rhat, b$diag$divergences))
```

```{r brms-table, echo = FALSE, eval = !is.null(N$brms)}
knitr::kable(N$brms$interaction, digits = 3,
  col.names = c("interaction term", "OR (median)", "OR 2.5%", "OR 97.5%", "P(OR > 1)"),
  caption = "Explicit gender x migration interaction, brms posterior")
```

The honest reading is **suggestive, not confirmed**. There is roughly a four-in-five
posterior chance that first-generation migrant women carry a NEET penalty beyond
additivity, but the interval crosses one, so it cannot be asserted. Crucially, that is
a *defensible* summary; the singular `lme4` zero was not. The interval is what `brms`
contributes, exactly as in the sparse-data vignette.

## Takeaways

* MAIHDA **revealed** the intersectional structure of European youth unemployment: a
  strong gradient (3--4&times; across strata) with real discriminatory accuracy
  (AUC `r sprintf("%.2f", U$auc)`), dominated by education and overlaid on a still-larger
  cross-country contextual effect.
* That inequality is **additive**, not multiplicative -- corroborated by the PCV, a
  shrinkage-free interaction test, and the raw rates. Additivity *is* an
  intersectionality finding; it adjudicates the additive-vs-multiplicative debate that
  the theory leaves open, and lands where most MAIHDA applications do.
* On NEET, where theory predicts a migrant-women interaction, the data give a hint that
  the variance decomposition can only summarise honestly with a **Bayesian interval** --
  a worked reminder not to read a singular `lme4` interaction of zero as evidence of
  none.

## Caveats and reproducibility

The ESS microdata is licence-restricted and is **not** distributed with the package;
only these derived summaries and figures are. The analysis is **unweighted** (ESS
design/post-stratification weights could be added via `engine = "wemix"`); pools three
rounds without a period term; covers `r m$n_countries` ESS countries (which include
non-EU members, so "Europe (ESS)" is not the EU-27); and rests on `r m$n_strata` strata,
few enough that the interaction *variance* is hard to estimate -- the very point of the
final section. To reproduce: download the ESS Round 9--11 integrated files into
`data-raw/ess/`, then run `data-raw/fetch_ess_youth.R` and
`data-raw/precompute_youth_vignette.R`.
```
