---
title: "Finding the interactions that matter "
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding the interactions that matter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Where is the intersectionality?

The headline VPC tells you *how much* of the variation lies between intersectional
strata; the PCV tells you how much of that is the *additive* sum of the dimensions'
main effects. Neither tells you **which strata** are genuinely intersectional --
whose outcome departs from what the additive parts predict. That departure is the
stratum random effect (BLUP) of the **adjusted** model, and `maihda_interactions()`
packages it into a flag of which strata are credibly non-zero.

```{r model}
library(MAIHDA)
data("maihda_health_data")

a <- maihda(
  BMI ~ Age + Gender + Race + Education + (1 | Gender:Race:Education),
  data = maihda_health_data
)
```

## `maihda_interactions()`

Pass the `maihda()` analysis and the function uses the **adjusted** model
automatically (that is the only model on which the stratum effect isolates the
*pure* interaction). It returns one row per stratum -- the interaction BLUP, its
interval, and whether it is flagged -- sorted flagged-first, then by magnitude:

```{r interactions}
ints <- maihda_interactions(a)
ints
```

The number of flagged strata is on the object's attributes:

```{r interactions-n}
sprintf("%d of %d strata flagged", attr(ints, "n_flagged"), attr(ints, "n_strata"))
```

Each row carries the `interaction` (BLUP, on the model's link scale), its
`lower`/`upper` interval, the stratum size `n`, the `direction` (above/below the
additive expectation), and -- for the frequentist engines -- the conditional `se`
and `p_value`.

### Screening many strata: FDR

With many strata you may want an explicit error-rate control. The default is
`adjust = "none"` (see the caveat below on why), but for a screen across many
strata the **false-discovery rate** (`adjust = "BH"`) matches the goal -- discovery
-- better than family-wise methods:

```{r interactions-bh}
maihda_interactions(a, adjust = "BH")
```

## Highlighting the flagged strata on the plots

Pass `highlight_interactions = TRUE` to `plot()` and the flagged strata are ringed
and starred on the BLUP-based views. On a `maihda()` analysis the flags are
computed once from the adjusted model and reused:

```{r plot-decomp, fig.height = 5}
plot(a, type = "effect_decomp", highlight_interactions = TRUE)
```

```{r plot-predicted, fig.height = 5}
plot(a, type = "predicted", highlight_interactions = TRUE)
```

You can also pass a precomputed `maihda_interactions` object (to reuse a specific
`conf_level`/`adjust`) instead of `TRUE`.

## How to read it -- and what not to conclude

- **It must come from the adjusted model.** On a null model the stratum effect is
  the *total* between-stratum deviation (additive + interaction), so passing a bare
  null model is flagged with a warning. Passing the `maihda()` analysis handles this
  for you.
- **It is on the link (latent) scale.** For a logistic model the interaction is a
  log-odds departure; the additive/interaction split is only exact on that scale.
- **Treat the flags as exploratory, not confirmatory.** The BLUP is already a
  shrunken (partially pooled) estimate, which protects against spurious extremes,
  so the relevant risk is sign/magnitude error rather than family-wise Type I error.
  That is *why* `adjust = "none"` is the default; corrections applied to shrunken
  estimates are conservative. Use `"BH"` when you need a discovery-rate story, and
  read the flagged strata as hypotheses to probe, not confirmed interactions.
- **A singular fit voids the flags.** When the between-stratum variance is pinned at
  the boundary the BLUP SEs are unreliable; the object records `attr(., "singular")`
  and "not flagged" is then not evidence of "no interaction" (see the
  [Bayesian sparse vignette](bayesian_sparse_maihda.html)).

## Bayesian evidence (brms)

For a `brms` fit the function uses the **exact posterior tail** rather than a normal
approximation: a credible interval and the probability of direction
`pd = P(BLUP > 0)`. `adjust` is inert (the Bayesian answer is multiplicity-free).

```{r brms, eval = FALSE}
a_brms <- maihda(BMI ~ Age + Gender + Race + Education + (1 | Gender:Race:Education),
                 data = maihda_health_data, engine = "brms")
maihda_interactions(a_brms)   # gains a `pd` column; no p-values
```

## See also

- [Reporting MAIHDA results](reporting_results.html) -- tidy output and publication tables.
- [Interpreting MAIHDA plots and diagnostics](interpreting_plots.html) -- the
  effect-decomposition and predicted views in depth.
- [Bayesian MAIHDA for sparse intersections](bayesian_sparse_maihda.html) -- when ML
  cannot see a real interaction.

## References

- Evans, C. R., Williams, D. R., Onnela, J. P., & Subramanian, S. V. (2018). A
  multilevel approach to modeling health inequalities at the intersection of multiple
  social identities. *Social Science & Medicine*, 203, 64-73.
- Gelman, A., Hill, J., & Yajima, M. (2012). Why we (usually) don't have to worry
  about multiple comparisons. *Journal of Research on Educational Effectiveness*, 5(2),
  189-211.
