---
title: "Finding interaction patterns"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding interaction patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Overview

One question that does come up naturaly is which strata depart most clearly
from the additive expectation?

For these exploratory diagnostics, `maihda()` can compute the stratum random effects and intervals from the adjusted model, and apply a multiplicity rule to flag strata that depart from the additive expectation more than expected by chance. The diagnostic is stored in the fitted model object and can be used to highlight strata in the effect-decomposition and predicted-strata plots.

## Run a standard analysis and choose the multiplicity rule

The default is `adjust = "none"`. This reports each stratum-level interval and
flag without a multiple-testing correction.

If the goal is to scan all strata and highlight a smaller set for follow-up, use
an adjustment such as Benjamini-Hochberg:

```{r model-bh}
library(MAIHDA)
model_bh <- maihda(
  BMI ~ Age + Gender + Race + Education + (1 | Gender:Race:Education),
  data = maihda_health_data,
  interactions = "BH" # Benjamini-Hochberg adjustment
)
```

The printed output reports how many strata were flagged and which adjustment rule
was used. The full table is stored in `a$interactions`.

```{r table}
model_bh$interactions
```

Each row is one stratum. The main columns are:

- `interaction`: the adjusted-model stratum random effect, on the model scale.
- `lower` and `upper`: the interval for that random effect.
- `direction`: whether the stratum is above or below the additive expectation.
- `flagged`: whether the stratum passes the selected screening rule.

For frequentist fits, the table also includes the conditional standard error,
p-value, and adjusted p-value when a correction is requested.

## Highlight flagged strata

The plotting methods can reuse the stored diagnostic.

```{r plot-decomp, fig.height = 5}
plot(model_bh, type = "effect_decomp", highlight_interactions = TRUE)
```

To highlight the strata that survive a specific adjustment ( in this case Benjamini-Hochberg ), pass the adjusted model object that contains the diagnostic with the desired adjustment:

```{r plot-predicted, fig.height = 5}
plot(model_bh, type = "predicted", highlight_interactions = TRUE)
```


## See also

- [Introduction to MAIHDA](introduction.html), the main two-model workflow.
- [Interpreting MAIHDA plots and diagnostics](interpreting_plots.html), how to
  read the effect-decomposition and predicted-strata plots.
- [Bayesian MAIHDA for sparse intersections](bayesian_sparse_maihda.html), sparse
  cells and uncertainty in variance components.

## References

- Evans, C. R., Williams, D. R., Onnela, J. P., & Subramanian, S. V. (2018). A
  multilevel approach to modeling health inequalities at the intersection of
  multiple social identities. *Social Science & Medicine*, 203, 64-73.

- Merlo, J. (2018). Multilevel analysis of individual heterogeneity and
  discriminatory accuracy (MAIHDA) within an intersectional framework. *Social
  Science & Medicine*, 203, 74-80.
