---
title: "Comparing Intersectional Inequality Across Groups"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Intersectional Inequality Across Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Introduction

MAIHDA is often used to estimate how much outcome variation lies between
intersectional strata in a single population. The group comparison workflow asks
a related contextual question: is intersectional inequality larger in some
higher-level groups than in others?

Examples include comparing gender-by-class inequality across countries,
race-by-gender inequality across regions, or intersectional inequality across
survey waves. In the MAIHDA package, this is handled by `maihda()` with a
`group` argument or, when you want only the group comparison table, by
`compare_maihda_groups()`.

## Example data: countries, gender, and socioeconomic status

The package includes `maihda_country_data`, a teaching dataset based on a
balanced subset of OECD PISA 2018 data. Each row is a student, the outcome is a
mathematics score, and the intersectional strata are formed by `gender` and
`ses`.

The higher-level grouping variable is `country`, which lets us ask whether the
VPC/ICC for gender-by-SES inequality differs across countries.

```{r data}
library(MAIHDA)
data("maihda_country_data")

country_counts <- as.data.frame(table(maihda_country_data$country))
names(country_counts) <- c("country", "n")
country_counts

table(maihda_country_data$gender, maihda_country_data$ses)
```

## One-call workflow

Use `maihda()` when you want the usual model summary and the group comparison in
one object. The formula below defines six intersectional strata
(`gender:ses`) and asks for a separate VPC/ICC estimate within each country.

```{r one-call}
analysis <- maihda(
  math ~ 1 + (1 | gender:ses),
  data = maihda_country_data,
  group = "country"
)

analysis
```

The printed report includes the overall VPC/ICC first, then the country-level
comparison. The group comparison is also stored in `analysis$groups`, so it can
be inspected or saved as a regular data frame.

```{r group-table}
group_results <- as.data.frame(analysis$groups)
group_results[order(group_results$vpc, decreasing = TRUE),
              c("group", "n", "n_strata", "vpc", "var_between",
                "var_residual", "status")]
```

In this example all countries have 600 students and all six gender-by-SES strata
are populated. Higher VPC/ICC values indicate that a larger share of mathematics
score variation lies between the intersectional strata within that country.

## Visualizing the comparison

The group VPC plot orders countries by the estimated intersectional VPC/ICC. If
the comparison was run with `bootstrap = TRUE`, the same plot will include
per-country confidence intervals.

```{r group-vpc-plot}
plot(analysis, type = "group_vpc")
```

The variance-components view shows the same comparison as shares of the total
model variance. The orange segment is the between-stratum component, which is
the numerator of the VPC/ICC.

```{r group-components-plot}
plot(analysis, type = "group_components")
```

## Direct group comparison

If you only need the group comparison table and plots, call
`compare_maihda_groups()` directly. This fits the same stratified per-country
models used by `maihda(group = "country")`.

```{r direct-workflow, eval = FALSE}
group_cmp <- compare_maihda_groups(
  math ~ 1 + (1 | gender:ses),
  data = maihda_country_data,
  group = "country"
)

group_cmp
plot(group_cmp, type = "vpc")
plot(group_cmp, type = "components")
```

By default, `shared_strata = TRUE`. That means the `gender:ses` strata are
defined once on the pooled data, then reused in every country. A "female:Low"
stratum therefore refers to the same type of stratum in each country, which is
what makes the VPC/ICC estimates directly comparable.

Set `shared_strata = FALSE` only when you intentionally want each group to
rebuild its own strata. In that case, stratum identities are no longer shared
across groups, so interpretation should focus on within-group structure rather
than direct stratum-to-stratum comparison.

## Adding bootstrap intervals

For publication-oriented summaries, you can request per-group parametric
bootstrap confidence intervals with the `lme4` engine. This can take noticeably
longer because each group requires repeated model refits, so the example is not
run when the vignette is built.

```{r bootstrap-example, eval = FALSE}
group_cmp_boot <- compare_maihda_groups(
  math ~ 1 + (1 | gender:ses),
  data = maihda_country_data,
  group = "country",
  bootstrap = TRUE,
  n_boot = 500
)

plot(group_cmp_boot, type = "vpc")
```

## Interpretation cautions

The group comparison is a stratified analysis: the package fits one MAIHDA model
within each country and compares the resulting VPC/ICC estimates. It is not a
cross-classified model, and it does not estimate country effects in a single
pooled multilevel model.

The bundled PISA data are for teaching and software demonstration only. They use
a balanced subsample, one plausible value for each achievement score, and no
survey weights or complex-sampling adjustments. Treat the results as a
walkthrough of the workflow, not as substantive cross-national evidence.
