---
title: "Comparing Intersectional Inequality Across Groups"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Intersectional Inequality Across Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Introduction

MAIHDA is often used to estimate how much outcome variation lies between
intersectional strata in a single population. The group comparison workflow asks
a related contextual question: is intersectional inequality larger in some
higher-level groups than in others?

Examples include comparing gender-by-class inequality across countries,
race-by-gender inequality across regions, or intersectional inequality across
survey waves. In the MAIHDA package, this is handled by `maihda()` with a
`group` argument or, when you want only the group comparison table, by
`compare_maihda_groups()`.

## Example data: countries, gender, and socioeconomic status

The package includes `maihda_country_data`, a teaching dataset based on a
balanced subset of OECD PISA 2018 data. Each row is a student, the outcome is a
mathematics score, and the intersectional strata are formed by `gender` and
`ses`.

The higher-level grouping variable is `country`, which lets us ask whether the
VPC/ICC for gender-by-SES inequality differs across countries.

```{r data}
library(MAIHDA)
data("maihda_country_data")

country_counts <- as.data.frame(table(maihda_country_data$country))
names(country_counts) <- c("country", "n")
country_counts

table(maihda_country_data$gender, maihda_country_data$ses)
```

## One-call workflow

Use `maihda()` when you want the full decomposition and the group comparison in
one object. The formula below defines six intersectional strata
(`gender:ses`) and asks for a separate decomposition within each country. Because
`maihda()` is a two-model workflow, it fits both the null and the adjusted model
overall *and* within each country, so the group table also carries a per-country
**PCV** (the additive share of that country's intersectional inequality).

```{r one-call}
# gender + ses are written as additive fixed effects (the adjusted model); maihda()
# derives the null by dropping them, both overall and within each country.
analysis <- maihda(
  math ~ gender + ses + (1 | gender:ses),
  data = maihda_country_data,
  group = "country"
)

analysis
```

The printed report includes the overall VPC/ICC and PCV first, then the
country-level comparison. The group comparison is also stored in `analysis$groups`,
so it can be inspected or saved as a regular data frame.

```{r group-table}
group_results <- as.data.frame(analysis$groups)
group_results[order(group_results$vpc, decreasing = TRUE),
              c("group", "n", "n_strata", "vpc", "var_between",
                "var_residual", "pcv", "status")]
```

In this example all countries have 600 students and all six gender-by-SES strata
are populated. Higher VPC/ICC values indicate that a larger share of mathematics
score variation lies between the intersectional strata within that country; the
`pcv` column shows how much of each country's between-stratum variance is the
additive sum of the gender and SES main effects (so a lower PCV flags inequality
that is more intersection-specific). Here the `pcv` is essentially 1 in every
country -- a real feature of this data, not an artefact; the PCV section below
explains why that flat result is itself the finding.

## Visualizing the comparison

The group VPC plot orders countries by the estimated intersectional VPC/ICC. If
the comparison was run with `bootstrap = TRUE`, the same plot will include
per-country confidence intervals.

```{r group-vpc-plot}
plot(analysis, type = "group_vpc")
```

The variance-components view shows the same comparison as shares of the total
model variance. The orange segment is the between-stratum component, which is
the numerator of the VPC/ICC.

```{r group-components-plot}
plot(analysis, type = "group_components")
```

## Share versus magnitude

The VPC/ICC is a *ratio*: the share of the unexplained variation that lies between
strata. A larger VPC therefore does not necessarily mean a larger *amount* of
between-stratum (intersectional) variation -- it can also reflect a smaller residual
variance. Two countries with the same absolute between-stratum variance can have very
different VPCs, and vice versa.

The `"group_between_variance"` view plots the absolute between-stratum variance
(`var_between`) directly, so it can be read alongside the VPC to separate the *share*
of inequality from its *magnitude*.

```{r group-between-variance-plot}
plot(analysis, type = "group_between_variance")
```

For non-Gaussian models this variance is on the model (link) scale, and unlike the
VPC it is not normalised by the residual variance.

## Additive share (PCV) by group

The `"group_pcv"` view bars each country's PCV -- the proportional drop in
between-stratum variance once the gender and SES main effects are added. A high
bar means that country's intersectional inequality is largely the additive sum of
its parts; a low (or negative) bar flags inequality that is more
intersection-specific. (With strata defined by a single dimension there is no
intersection to decompose, so this view and the `pcv` column are unavailable.)

In this real-data example the view is deliberately anticlimactic: every bar sits
at (essentially) 100%. The PISA gender-by-SES achievement gaps are almost purely
additive, so once the two main effects enter the model no between-stratum
variance remains in any country -- each per-country adjusted fit lands on the
singular boundary, which is what the `boundary (singular) fit` messages emitted
while the models were fitted mean here. The flat row of full-height bars is
itself the substantive result: none of these six countries shows
intersection-specific gender-by-SES inequality in mathematics. The view earns its
keep when the bars *differ* -- a country with a visibly lower (or negative) bar
would be the one whose inequality is more than the sum of its parts.

```{r group-pcv-plot}
plot(analysis, type = "group_pcv")
```

## Direct group comparison

If you only need the group comparison table and plots, call
`compare_maihda_groups()` directly. This fits the same per-country null and
adjusted models used by `maihda(group = "country")` (the `pcv` column appears when
the strata have at least two dimensions).

```{r direct-workflow, eval = FALSE}
group_cmp <- compare_maihda_groups(
  math ~ 1 + (1 | gender:ses),
  data = maihda_country_data,
  pvc_
  group = "country"
)

group_cmp
plot(group_cmp, type = "vpc")
plot(group_cmp, type = "components")
plot(group_cmp, type = "pcv")
```

By default, `shared_strata = TRUE`. That means the `gender:ses` strata are
defined once on the pooled data, then reused in every country. A "female:Low"
stratum therefore refers to the same type of stratum in each country, which makes
the stratum *definitions* comparable across countries.

Shared definitions are necessary but not by themselves sufficient for the VPCs to
be *directly* comparable. A given country may not contain every stratum, so two
countries' VPCs can be estimated over different sets of *populated* strata. When
that happens, `compare_maihda_groups()` issues a warning, and the VPCs should be
read with that caveat in mind (alongside the `n_strata` column) rather than as
strictly like-for-like.

Set `shared_strata = FALSE` only when you intentionally want each group to
rebuild its own strata. In that case, stratum identities are no longer shared
across groups, so interpretation should focus on within-group structure rather
than direct stratum-to-stratum comparison.

## Adding bootstrap intervals

For publication-oriented summaries, you can request per-group parametric
bootstrap confidence intervals with the `lme4` engine. This can take noticeably
longer because each group requires repeated model refits, so the example is not
run when the vignette is built.

```{r bootstrap-example, eval = FALSE}
group_cmp_boot <- compare_maihda_groups(
  math ~ 1 + (1 | gender:ses),
  data = maihda_country_data,
  group = "country",
  bootstrap = TRUE,
  n_boot = 500
)

plot(group_cmp_boot, type = "vpc")
```
