---
title: "Youth MAIHDA: NEET and well-being across Europe"
author: "Hamid Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Youth MAIHDA: NEET and well-being across Europe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Why MAIHDA for youth research?

Youth-research institutes -- the German Youth Institute (DJI), the *Youth & Work*
programme and the *Observatoire de la jeunesse* in Luxembourg -- ask, again and
again, the same two questions about young people's life chances:

- the **school-to-work transition**: who ends up **NEET** (Not in Employment,
  Education or Training), the headline indicator of a stalled transition; and
- **subjective well-being**: how good young people feel their lives are.

And they ask these questions *intersectionally*. A young woman with a migration
background whose parents have little formal education is not simply "a woman" plus
"a migrant" plus "low social origin": the disadvantages can **compound** in ways
that looking at one axis at a time will miss. That is exactly what MAIHDA is built
for -- it places every young person in an **intersectional stratum** (here
`gender x migration x parental education`) and asks how much of the inequality in
an outcome lies *between* those strata, and how much of that is the simple additive
sum of the parts versus a genuine intersectional **interaction**.

This vignette walks through the full MAIHDA toolkit on a youth dataset:

1. the two-model decomposition (**VPC/ICC** and **PCV**) for a continuous outcome
   (well-being),
2. the **discriminatory-accuracy** workflow for a binary outcome (NEET),
3. comparing intersectional inequality **across countries**, and
4. pinpointing **which intersections** carry a genuine interaction.

## The data

`maihda_youth_data` is a simulated cross-section of 16--29-year-olds in four
countries. It is **synthetic** -- the real surveys it is modelled on (the Youth
Survey Luxembourg, the DJI's AID:A, and the youth sub-sample of the European Social
Survey) require data-use agreements and cannot be shipped in a package -- but its
design and its NEET / life-satisfaction levels are calibrated to Eurostat figures
for young Europeans, so the workflow below mirrors a real analysis.
j
> **Teaching data.** The values are constructed, not observed; read the patterns
> below as an illustration of the *method*, not as findings about these countries.

```{r load}
library(MAIHDA)
data("maihda_youth_data")

str(maihda_youth_data)
```

The three stratum dimensions form `2 x 2 x 3 = 12` intersectional strata; `country`
is a higher-level grouping variable. Every stratum is well populated (this is not a
sparse-cell problem -- see `vignette("bayesian_sparse_maihda")` for that case):

```{r strata}
# the 12 intersectional strata and their sizes (pooled across countries)
table(maihda_youth_data$gender,
      maihda_youth_data$migration,
      maihda_youth_data$parental_edu)

# NEET prevalence differs across countries, as in the real Eurostat series
round(tapply(maihda_youth_data$neet == "Yes", maihda_youth_data$country, mean), 3)
```

## Well-being: the two-model decomposition

`maihda()` runs the standard pipeline in one call. Writing the stratum dimensions'
main effects in the formula tells `maihda()` they are the **adjusted** model; it
derives the **null** model (the intersectional random intercept alone) by dropping
them, fits both on the same sample, and reports the VPC/ICC and the PCV.

```{r wb-fit}
wb <- maihda(
  wellbeing ~ gender + migration + parental_edu +
    (1 | gender:migration:parental_edu),
  data = maihda_youth_data
)
wb
```

The **VPC/ICC** is the share of well-being variation that lies *between* the
intersectional strata. The **PCV** (proportional change in variance) is how much
of that between-stratum variance the additive main effects account for: a high PCV
means inequality is largely the additive sum of gender, migration, and social
origin; the remainder is the part the additive model cannot reach -- the candidate
**intersectional** signal. (As always, the PCV is a model-dependent variance change,
not a causal decomposition.)

```{r wb-summary}
summary(wb)
wb$pcv
```

`plot()` sends each view to the model it belongs on automatically -- the VPC and
shrinkage views to the null model, the additive-vs-intersectional views to the
adjusted model.

```{r wb-plot-vpc}
plot(wb, type = "vpc")              # variance partition (null model)
```

```{r wb-plot-effect-decomp}
plot(wb, type = "effect_decomp")   # additive vs intersectional (adjusted model)
```

```{r wb-plot-predicted}
plot(wb, type = "predicted")       # predicted well-being per stratum, with 95% CIs
```

The predicted-value plot ranks the 12 strata: youth with a migration background and
low parental education sit at the bottom, those without and with highly educated
parents at the top -- the additive gradient. The effect-decomposition plot is where
the *intersectional* part becomes visible, separated from that additive gradient.

## NEET: discriminatory accuracy

NEET is binary, so MAIHDA fits a multilevel **logistic** model. `maihda()`
auto-detects the two-level outcome (it reports how it recoded the factor); passing
`family = "binomial"` makes it explicit. For a binary outcome the **discriminatory
accuracy** -- the AUC / C-statistic and the Median Odds Ratio (MOR) -- rides along
on the summary automatically.

```{r neet-fit}
neet <- maihda(
  neet ~ gender + migration + parental_edu +
    (1 | gender:migration:parental_edu),
  data = maihda_youth_data, family = "binomial"
)
neet
```

Two numbers tell complementary stories. The **VPC** says a substantial share of the
(latent) variation in NEET risk lies between intersectional strata. The **AUC**, by
contrast, asks a prediction question: *given only which intersectional stratum a
young person belongs to, how well can we separate those who are NEET from those who
are not?* Here the AUC is moderate -- and that is the central, cautionary message of
the "DA" in MAIHDA: a real between-stratum gradient still translates into only
limited accuracy at the **individual** level, so intersectional strata describe
*structure* far better than they *target individuals*. You can pull the pieces out
directly:

```{r neet-da}
maihda_discriminatory_accuracy(neet$model)   # strata-only (null) model: Merlo's DA
```

The VPC for a logistic model is on the **latent** scale (the level-1 variance is
fixed at `pi^2/3`); see `vignette("binary_outcomes")` for the full treatment,
including the probability-scale complement `maihda_vpc_response()`.

```{r neet-plot}
plot(neet, type = "predicted")     # predicted NEET probability per stratum
```

## Comparing inequality across countries

Add `group = "country"` and `maihda()` runs the whole decomposition *within each
country* as well, so we can ask the contextual question youth institutes care
about: **is intersectional inequality sharper in some countries than others?**

```{r group-fit, message = FALSE, warning = FALSE}
by_country <- maihda(
  neet ~ gender + migration + parental_edu +
    (1 | gender:migration:parental_edu),
  data = maihda_youth_data, group = "country", family = "binomial"
)

group_results <- as.data.frame(by_country$groups)
group_results[order(group_results$vpc, decreasing = TRUE),
              c("group", "n", "n_strata", "vpc", "var_between", "pcv", "status")]
```

```{r group-plot-vpc}
plot(by_country, type = "group_vpc")
```

The VPC/ICC is a *share*, so a country can post a high VPC partly because its
residual variation is small. Read it alongside the **absolute** between-stratum
variance, which is the magnitude of intersectional inequality rather than its share:

```{r group-plot-magnitude}
plot(by_country, type = "group_between_variance")
```

## Which intersections actually interact?

A high PCV says most inequality is additive -- but *some* is not. `maihda_interactions()`
reads each stratum's interaction effect (its departure from the additive
prediction, the random effect of the adjusted model) and flags those credibly
different from zero. With many strata, screen with an FDR correction:

```{r interactions}
flagged <- maihda_interactions(wb, adjust = "BH")
flagged
```

The flagged strata are not scattered at random: they are the `migration x social
origin` cells. This is what an intersectional interaction *is* -- the effect of one
dimension **depends on** another. Here the social-origin gradient in well-being is
**steeper among youth with a migration background**: low parental education carries
an extra penalty when combined with a migration background, beyond what either
factor predicts on its own. Equivalently, the additive model *over-credits* high
parental education for youth without a migration background. That dependence -- not a
single "worst-off" cell -- is the intersectional finding, and it is exactly the kind
of compounding disadvantage the additive main effects cannot represent.

> **What an interaction can and cannot look like.** A genuine interaction (one that
> survives adjustment for the main effects) is *orthogonal* to them, so it always
> moves cells in a symmetric pattern -- "the doubly-disadvantaged get worse *and* the
> doubly-advantaged get better" is the signature of an **additive** effect, which
> the main effects already absorb. Read the flagged strata as *where the additive
> model mis-predicts*, and in which direction.

## Reading this for youth policy

Put together, the workflow gives a youth-policy reader four distinct things:

- **VPC/ICC** -- how segregated outcomes are across intersectional groups (a
  structural-inequality summary);
- **PCV** -- how much of that is the simple additive sum of the axes versus genuine
  intersectionality (whether single-axis policies would suffice);
- **discriminatory accuracy** -- whether intersectional membership is sharp enough to
  *target* individuals, or only to *describe* groups (a caution against profiling);
- **the group comparison** -- where intersectional inequality is most acute, to
  prioritise contexts.

For the modelling extensions most relevant to youth data -- survey weights
(`sampling_weights=`), youth nested in schools or regions (`context=`), panel
/ growth-curve designs (`vignette("longitudinal")`), and small intersectional cells
(`vignette("bayesian_sparse_maihda")`) -- see the linked workflows.

## References

- Merlo, J. (2018). Multilevel analysis of individual heterogeneity and
  discriminatory accuracy (MAIHDA) within an intersectional framework.
  *Social Science & Medicine*, 203, 74-80.

- Evans, C. R., Williams, D. R., Onnela, J. P., & Subramanian, S. V. (2018). A
  multilevel approach to modeling health inequalities at the intersection of
  multiple social identities. *Social Science & Medicine*, 203, 64-73.

- Merlo, J., Wagner, P., Ghith, N., & Leckie, G. (2016). An original stepwise
  multilevel logistic regression analysis of discriminatory accuracy: the case of
  neighbourhoods and health. *PLOS ONE*, 11(4), e0153778.

- Crenshaw, K. (1989). Demarginalizing the intersection of race and sex.
  *University of Chicago Legal Forum*, 1989(1), 139-167.
